---
title: "Data Visualization Project"
author: "Nate Hepler"
date: "6/19/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)

## set working directory
knitr::opts_knit$set(root.dir = '/cloud/project')
```

# Part 1: Exploratory Data Analysis

The data we'll be using for this part of the project were downloaded from [kaggle](https://www.kaggle.com/aaronschlegel/seattle-pet-licenses/data), and include information about "pet licenses issued by the Seattle Animal Shelter between 2005 and early 2017." We'll be exploring these data and generating a few exploratory plots in the first part of the project.


## The Data

First, we'll read the data in from our `data/raw_data` directory.

```{r pet-data, warning=FALSE} 
pets <- read.csv("dataviz_project/data/raw_data/seattle_pet_licenses.csv", stringsAsFactors = FALSE)
```

```{r pet-explore, warning=FALSE} 
## Include code here to explore the data 
## and answer the questions in the exercise
```

## Visualize the Data
### Breaking down by animal
Deteriming how many animals of each three species have a license
```{r pet-barchart, warning=FALSE} 
## visual breakdown of how many of each species 
## are included in the dataset
ggplot(pets) +
  geom_bar(aes(x=species, fill=species)) + 
  ylab('Number of animals') + 
  xlab('Species') + 
  labs(title = 'Number of animals by species', 
       subtitle='Three different species: Cat, Dog and Livestock') +
  theme(axis.text = element_text(color='black'),
        legend.position = 'none')
```

### Top 10 Cat Names
```{r pet-cat-table, warning=FALSE} 
## Table: Most frequent Cat Name 
pets %>% 
  filter(species == "Cat", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Cat Names in Seattle")
```

### Top 10 Dog Names
```{r pet-dog-table, warning=FALSE} 
## Table: Most frequent Dog Name
pets %>%
  filter(species == 'Dog', animal_s_name != '') %>%
  group_by(animal_s_name) %>%
  summarise(n=n()) %>%
  arrange(-n) %>%
  top_n(n=10) %>%
  knitr::kable(., caption = 'Top 10 Dog Names in Seattle')
```

### Calculating and graphing licenses issued over time
```{r pet-linechart, warning=FALSE} 
## add date and ym columns
pets$date <- lubridate::ymd_hms(pets$license_issue_date)
pets$ym <- as.yearmon(pets$date, "%y%m")

## how the number of licenses recorded has changed over time
pets %>%
  ## group by yearmonth (`ym`)
  group_by(ym) %>% 
  ## count number within each group
  summarise(n=n()) %>%
  ggplot(., aes(x=ym, y=n)) + 
  ## geom name for line chart
  geom_line() +
  scale_x_yearmon() + 
  xlab("") + 
  ylab("Number of licenses")
```

### Graphing number of licenses by species
```{r pet-plot, warning=FALSE} 
### Explore the data -- create your own plot!
pets %>%
  ## group by yearmonth (`ym`)
  group_by(ym, species) %>% 
  ## count number within each group
  summarise(n=n()) %>%
  ggplot(., aes(x=ym, y=n, color=species)) +
  geom_line() + 
  scale_x_yearmon() + 
  xlab('') +
  ylab('Number of licenses')
  
```

# Part 2: Explanatory Data Analysis

The data used in this part of the project were dowloaded from [FiveThirtyEight](https://data.fivethirtyeight.com/) - `steak-survey`.  They were originally used in the article: [How Americans Like Their Steak](https://fivethirtyeight.com/features/how-americans-like-their-steak/). The goal of this part of the project will be to recreate the data visualization used in this article.

## The Data

```{r meat-data, warning=FALSE} 
## read in the data
steak <- read.csv("dataviz_project/data/raw_data/steak-risk-survey.csv", stringsAsFactors = FALSE) %>%
  ## remove first row which just includes the word "Response" in each column
  slice(2:n())
```

## Explore the Data

```{r meat-explore, warning=FALSE} 
## Include code here to explore the data
steak_explore <- steak %>%
  group_by(Consider.the.following.hypothetical.situations...br.In.Lottery.A..you.have.a.50..chance.of.success..with.a.payout.of..100...br.In.Lottery.B..you.have.a.90..chance.of.success..with.a.payout.of..20...br..br.Assuming.you.have..10.to.bet..would.you.play.Lottery.A.or.Lottery.B.) %>%
  summarise(n=n())

steak_yes <- steak %>% 
  group_by(Do.you.eat.steak.) %>%
  summarise(n=n())

steak_prep <- unique(steak$How.do.you.like.your.steak.prepared.)
steak_prep2 <- steak %>%
  group_by(How.do.you.like.your.steak.prepared.) %>%
  summarise(n=n())

  
## and answer the questions in the exercise

```

## Wrangle the Data

Simplifying data to just include steak preferences.

```{r meat-wrangle, warning=FALSE} 
## wrangle the steak data set
pref <- steak %>% 
  mutate(steak_pref = factor(How.do.you.like.your.steak.prepared.,
                             levels = c("Well",
                                        "Medium Well",
                                        "Medium",
                                        "Medium rare",
                                        "Rare"))) %>% 
  filter(steak_pref != "") %>% 
  group_by(steak_pref) %>% 
  summarise(n = n()) %>%
  mutate(prop = n / sum(n))
```

## Visualize the Data

Creating bar chart for steak preference
```{r meat-visualize, warning=FALSE}
## generate the plot 
p <- ggplot(pref) +
  ## specify you want to generate a bar chart
  geom_bar(aes(x = steak_pref, y = prop, fill = steak_pref),
                stat = 'identity', 
                width = 0.7) + 
  ## this adds text labels (you don't have to change anything here)
  geom_text(aes(label = paste0(as.integer(prop*100),"%"),
                x = steak_pref,
                y = prop), 
            stat= "identity", 
            hjust = -0.2, 
            size = 5,
            color = "grey40") +
  ## flip coordinates to make horizontal box plot
  coord_flip() +
  ## change the colors of the bars
  scale_fill_manual(values = c("#370B01", 
                           "#70402C",
                           "#955732", 
                           "#B36153",
                           "#B5413F")) +
  ## change the scale/labels of the steak-wellness variable (x-axis)
  scale_x_discrete(limits = levels(fct_rev(pref$steak_pref)),
                labels = c("Well", 
                           "Medium Well" = "Medium-well",
                           "Medium",
                           "Medium rare"="Medium-rare",
                           "Rare")) +
  ## change the scale/lables of the percent axis (y-axis)
  scale_y_continuous(labels = scales::percent, 
                expand = c(mult = c(0,0), 
                           add = c(0,0.035))) +
  ## change the title, subtitle, and caption
  labs(title="'How Do You Like Your Steak Prepared?'", 
                subtitle="From a survey of 432 steak-eating Americans",
                caption="FiveThirtyEight: Surveymonkey") +
  ## change the theme (use ggthemes)
  ggthemes::theme_fivethirtyeight() +
  ## fine tune the theme
  theme(axis.text = element_text(size = 14), 
                title = element_text(size = 16), 
                legend.position="none", 
                plot.caption=element_text(size = 12), 
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.text.x = element_blank() 
  )

p
```

## Save the Plot

```{r save-plot, warning=FALSE}
## save plot to figures/explanatory_figures directory
ggsave(plot = p, filename = 'dataviz_project/figures/explanatory_figures/steak_R.png', width = 8, height = 4)
```

## Visualize the Data

Wrangling data by smoker v non-smoker and then graphing.
```{r meat-plot, warning=FALSE} 
### Explore the data -- create your own plot!
smoke <- steak %>%
  mutate(smoke_choice = factor(Do.you.ever.smoke.cigarettes., 
                               levels = c('Yes', 'No'))) %>%
  filter(smoke_choice != '') %>%
  group_by(smoke_choice) %>%
  summarise(n=n()) %>%
  mutate(prop = n / sum(n))

p_smoke <- ggplot(smoke) +
  geom_bar(mapping = aes(x=smoke_choice, y=prop, fill = smoke_choice), 
           stat='identity', 
           width = 0.5) + 
  geom_text(mapping = aes(label = paste0(as.integer(prop*100), '%'), 
            x=smoke_choice, 
            y=prop),
            hjust = -0.1,
            size = 4) + 
  ggthemes::theme_fivethirtyeight() + 
  scale_fill_manual(values = c('navajowhite3', 'orange4')) + 
  coord_flip() + 
  scale_y_continuous(labels = scales::percent, 
                expand = c(mult = c(0,0), 
                           add = c(0,0.035))) + 
  labs(title = "'Do you ever smoke cigarettes'",
       subtitle = 'From a survey of 537 Americans',
       caption = 'FiveThirtyEight: Surveymonkey') + 
  theme(
    plot.title = element_text(size = 20),
    plot.subtitle = element_text(size = 16),
    plot.caption = element_text(size = 12),
    legend.position = 'none',
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size=15)
  )
  
p_smoke

```

## Save this plot 

```{r save-meat-plot, warning=FALSE}
## save plot to figures/explanatory_figures directory
ggsave(plot = p_smoke, filename = 'dataviz_project/figures/explanatory_figures/smoke_R.png', width=10, height=4)
```
